### 2018-04-20 concepts.tex
ok

In languages with a clear mechanism for expressing failure of a call
(e.g., exceptions in Python and Java, or Option/Result types in Rust,
Haskell, and ML), failure determinism can be automatically checked.
Moreover, the expected overhead should be even lower than the general
overhead for vertical determinism, in that we can expect most failing
operations to be fast in an automated test generation setting (that the
parameters to a call are illegal is the most common cause for
failure).

@@

>>>>

In languages with a clear mechanism for expressing failure of a call
(e.g., exceptions in Python and Java, or Option/Result types in Rust,
Haskell, and ML), failure determinism can be automatically checked.
Moreover, the expected overhead should be even lower than the general
overhead for vertical determinism, in that we can expect most failing
operations to be fast (since in test generation, failure is usually
due to invalid parameters).

@@

Finally, we note that this problem is not limited to debugging
nondeterminism/flakiness itself.  In some settings, we have observed
delta-debugging taking a test that had such a high probability of
failure we were unaware it was ``flaky'' (large runs showed there was
indeed a small chance of the test passing) and transforming it into a
test that only failed once in every 10 to 20 executions.
Delta-debugging, without taking potential nondeterminism into account,
can act as a flakiness magnifier, with unfortunate consequences for
debugging.  Multiple evaluations, with replications, are our proposed solution.

### 2018-04-20 experiments.tex
ok

Using TSTL's harness for {\tt redis-py}, both {\tt redis-py} and Redis
itself can be tested; unfortunately, generating stand-alone
high-coverage regression tests for {\tt redis-py} and Redis using TSTL
has proven difficult, as numerous Redis commands introduce nondeterministic
behavior.   Using {\tt tstl\_afl\_fuzz}, we can see that the {\tt redis-py} has a 
stability of only 56.26\%, a clear indicator of nondeterminism. Some
of the problematic commands are obvious simply by inspection (e.g.,
{\tt randomkey, srandmember}).  And, given an understanding of Redis
semantics, it is also clear that the various
commands producing data with a limited lifetime (e.g., {\tt expire,
  pexpire}) also introduce timing-based nondeterminism.

>>>>

Using TSTL's harness for {\tt redis-py}, both {\tt redis-py} and Redis
itself can be tested; unfortunately, generating stand-alone
high-coverage regression tests for {\tt redis-py} and Redis using TSTL
has proven difficult, as numerous Redis commands introduce nondeterministic
behavior:  thus the resulting tests are very often \emph{flaky}.   Using {\tt tstl\_afl\_fuzz}, we can see that the {\tt redis-py} has a 
stability of only 56.26\%, a clear indicator of nondeterminism. Some
of the problematic commands are obvious simply by inspection (e.g.,
{\tt randomkey, srandmember}).  And, given an understanding of Redis
semantics, it is also clear that the various
commands producing data with a limited lifetime (e.g., {\tt expire,
  pexpire}) also introduce timing-based nondeterminism.

### 2018-04-20 implementation.tex
ok

Another example of process-based nondeterminism was discovered by numerous Python developers when Python
version 3.3 introduced automatic random salting of hashes on a
per-process basis (a new salt chosen for each Python process),
in order to mitigate hash-based denial of service attacks
\cite{denial}.  Until version 3.6 this not only resulted in changes in
exact hash values, but in the order of iteration on
dictionaries\footnote{In version 3.6, the default dictionary
  implementation became an ordered dictionary with consistent
  iteration, though actual hashes remained nondeterministic.}.  While
relatively few programs rely on exact hash values, many implicitly
relied on them in that they only functioned correctly with a
predictable order for dictionary iteration.  Testing
Python code for nondeterminism based on this hash seed requires
running in a new process:  Python does not allow changing the salt, since changing hash values on-the-fly would
break all existing dictionaries and other structures relying on
hashing.

@@

\caption{TSTL Method Calls for Nondeterminism Detection}
\label{tab:methods}
\begin{tabular}{l|l}
%\hline
Name & Purpose \\
\hline
\hline
{\tt nondeterministic} & returns True if test exhibits final state nondeterminism in $k$
  tries with delay $d$\\
\hline
{\tt stepNondeterministic} & as above, except checks all visible values
                       (compares state at every step)\\
\hline
{\tt processNondeterministic} & as above, except runs test provided in
                          subprocess and compares non-opaque output
                          values \\
\hline
{\tt P} & returns probability that a predicate holds for a test (with
          optional number of sample to use in estimate) \\
\hline
{\tt forceP} & wraps a predicate for test reduction, checking that it
               has desired probability of holding, using
         sample size and replications parameters \\
\hline
\hline

@@

We implemented our approach in the TSTL \cite{NFM15,ISSTA15,tstlsttt} system, a
language and tool for property-based testing \cite{Hypothesis,ClaessenH00} of Python code. TSTL has
been used to detect (and usually fix) errors in a number of widely
used Python libraries, the Python implementation itself, the Solidity
compiler and a Solidity static analysis tool, and Mac OS \cite{tstl}.
Because TSTL supports differential testing \cite{tstlsttt}, horizontal
nondeterminism detection can technically be implemented simply by
declaring a system to be its own reference, using TSTL's notation for
differential testing.  However, such an approach requires some effort
on the part of the user to control which values are checked for
equivalence, and does not easily support injecting timing differences,
or using a new process, in checking the behavior.

We therefore instead made nondeterminism detection a first-class
property in TSTL, using TSTL's existing notation for marking some
types of values as \emph{opaque} (not usefully compared for equality),
used in automatic abstraction-based testing.  In TSTL, a test is a
sequence of \emph{actions}; parameter values to method and function
calls of the Software Under Test (SUT) are stored in finite-sized variable \emph{pools}.  A
TSTL test is (essentially) a sequence of assignments to pool values,
and method/function calls using pool values (including method calls on
pool objects).  This formalism is a widely used and efficient way to
represent unit tests \cite{Pacheco,AndrewsTR}.  In TSTL, \emph{visible
value determinism} is based on comparing the values of \emph{all pool
variables} after each test action (pool values other than the one
assigned to must also be compared, since a common source of
nondeterminism is when a call results in a change to a value passed as
a parameter, or even due to a shared reference held by two values).
TSTL omits comparison only of pools marked as {\tt OPAQUE} in the
TSTL language test harness (and values that do not support equality
checking).  \emph{Final State Determinism} simply performs the same
comparison, but only on the final values of all pool variables after a test has
finished executing.

>>>>

Another example of process-based nondeterminism was discovered by numerous Python developers when Python
version 3.3 introduced automatic random salting of hashes on a
per-process basis (a new salt chosen for each Python process),
in order to mitigate hash-based denial of service attacks
\cite{denial}.  Until version 3.6 this not only resulted in changes in
exact hash values, but in the order of iteration on
dictionaries\footnote{In 3.6, the default dictionary
  implementation became an ordered dictionary with consistent
  iteration, though actual hashes remained nondeterministic.}.  While
relatively few programs rely on exact hash values, many implicitly
relied on them in that they only functioned correctly with a
predictable order for dictionary iteration.  Testing
Python code for nondeterminism based on this hash seed requires
running in a new process:  Python does not allow changing the salt, since changing hash values on-the-fly would
break all existing dictionaries and other structures relying on
hashing.

@@

\caption{TSTL Method Calls for Nondeterminism Detection}
\label{tab:methods}
\begin{tabular}{l|l}
%\hline
Name & Purpose \\
\hline
\hline
{\tt nondeterministic} & returns True if test exhibits final state nondeterminism in $k$
  tries with delay $d$, optionally only over pools in $\mathcal{P}$\\
\hline
{\tt stepNondeterministic} & as above, except checks all visible values
                       (compares state at every step)\\
\hline
{\tt processNondeterministic} & as above, except runs test provided in
                          subprocess and compares non-opaque output
                          values \\
\hline
{\tt P} & returns probability that a predicate holds for a test (with
          optional number of sample to use in estimate) \\
\hline
{\tt forceP} & wraps a predicate for test reduction, checking that it
               has desired probability of holding, using
         sample size and replications parameters \\
\hline
\hline

@@

We implemented our approach in the TSTL \cite{NFM15,ISSTA15,tstlsttt} system, a
language and tool for property-based testing \cite{Hypothesis,ClaessenH00} of Python code. TSTL has
been used to detect (and usually fix) errors in a number of widely
used Python libraries, the Python implementation itself, the Solidity
compiler and a Solidity static analysis tool, and Mac OS \cite{tstl}.
Because TSTL supports differential testing \cite{tstlsttt}, horizontal
nondeterminism detection can technically be implemented simply by
declaring a system to be its own reference, using TSTL's notation for
differential testing.  However, such an approach requires considerable effort
on the part of the user to express which values are checked for
equivalence, and does not (without a great deal of effort) support injecting timing differences,
or re-executing in a new process, in checking for nondeterminism.

We therefore instead made nondeterminism detection a first-class
property in TSTL, using TSTL's existing notation for marking some
types of values as \emph{opaque} (not usefully compared for equality),
used in automatic abstraction-based testing.  In TSTL, a test is a
sequence of \emph{actions}; parameter values to method and function
calls of the Software Under Test (SUT) are stored in finite-sized variable \emph{pools}.  A
TSTL test is (essentially) a sequence of assignments to pool values,
and method/function calls using pool values (including method calls on
pool objects).  This formalism is a widely used and efficient way to
represent unit tests \cite{Pacheco,AndrewsTR}.  In TSTL, \emph{visible
value determinism} is based on comparing the values of \emph{all pool
variables} after each test action (pool values other than the one
assigned to must also be compared, since a common source of
nondeterminism is when a call results in a change to a value passed as
a parameter, or even due to a shared reference held by two values).
TSTL omits comparison only of pools marked as {\tt OPAQUE} in the
TSTL language test harness (and values that do not support equality
checking).  \emph{Final State Determinism} simply performs the same
comparison, but only on the final values of all pool variables (or a
set of designated pools) after a test has
finished executing.

### 2018-04-20 introduction.tex
ok

A system is deterministic if, given the system's complete state at a point in
time, it is possible, in principle, to predict its future behavior
perfectly.  We say ``in principle'' because in the real world,
prediction may be possible but hopelessly impractical.  We write
complex software systems in many cases because we cannot predict their
behavior (if we could perform the calculations in advance, ourselves,
we would just do so).  As a consequence, in software, rather than
defining determinism in terms of prediction, we usually therefore
simply say that a system is deterministic if, given the same state and
inputs, it always produces the same outputs\footnote{Of course, this
  shift of definition is often adopted in arguments for physical or
  human behavioral determinism, due to the even greater difficulties
  of prediction in those arenas.}.  The change of
definition, we will see, is central to our approach, and comes with a
number of limitations.

@@

Because the
programmer's abstraction of the system ignores  such a large number of
details, very few tests are in fact ``deterministic'' in the sense
that they eliminate all changes in behavior between executions.  Running the same test
twice almost always results in differences, given a low enough level of
abstraction, since the system load,
cache contents, branch predictor history, etc. are almost never
controlled for; however, this kind of
nondeterminism is usually of no interest, unless the test involves
extremely precise timing constraints.  Rather, what matters is when
some kind of nondeterminism unexpectedly impacts computed values at a
higher level of abstraction; in general, if the operating system
itself is not buggy, low-level nondeterminism, by design, is invisible
except in fine-grained performance testing or real-time systems.
Unexpected nondeterminism usually arises when there is an element of higher-level
state or input that \emph{is} critical to the produced behavior, but
the programmer has not anticipated.  E.g.,  when it is believed that the
behavior of a thread scheduler will not matter, but a race condition
in the code means that it does matter, or when the order of items in
an iterator on a hash table is important, and the hash values used are
salted and thus not predictable.

@@

In order to help ensure that they are reliable and secure, complex
modern software systems usually include a large set of
\emph{regression tests}.  A regression test suite is a (usually large)
set of tests that can be run against a software system every time it
is modified, to ensure that the modification has not broken the system
in some way.  \emph{Flaky tests} \cite{miccoflaky} are regression
tests that fail in an intermittent, unreliable fashion, and thus
degrade the utility of regression testing.  The essence of a flaky
test is that, for the same snapshot of test code and code under test
(CUT), it sometimes fails and sometimes passes: the pass/fail result (disposition) of the
test is not a deterministic property of the test code, code under
test, and environment running the test.  This produces three serious
problems: first, a flaky test often wastes developer time and delays
software changes by forcing the investigation of code-under-test that
is not actually incorrect.  Second, failures in flaky tests (for that
reason) are often ignored, and therefore serious software faults may
be missed.  Finally, to mitigate these problems, flaky tests are often
run multiple times, wasting valuable computing resources and also
delaying acceptance of code changes.  A canary in a coal mine is of
little use if canaries frequently become ill for reasons unrelated to
the presence of toxic gases.  Mining may stop for no good reason, or
miners may learn to ignore the canary, leading to tragedy; a third,
more ``practical'' option is that miners may carry so many redundant
canaries into the coal mine that canary-care becomes a serious burden
on mining.

@@

This paper proposes (1) a number of formal definitions of types of
nondeterminism and (2) a number of techniques, based on these definitions, for detecting and debugging
nondeterminism in property-based testing, particularly for library
code.  The methods also work for testing at the system level, but are
perhaps most effective for identifying the low-level sources of
nondeterminism in libraries.  This enables a number of useful results:
 the automated production of
nondeterminism-free regression tests even for libraries with some
sources of nondeterminism, the documentation of library
nondeterminism, and, most importantly, detection and effective debugging of
unexpected nondeterminism that is, itself, a bug.  We first
describe the types of nondeterminism we aim to detect, and the
extension of a widely used debugging algorithm to better handle
nondeterminism as a test property, then describe an implementation of
these ideas in a Python testing tool that is used on real-world
projects.  Finally, we use a set
of case studies of real-world Python libraries to demonstrate the
utility of our ideas and understand the cost of adding nondeterminism
detection to existing automated testing.

>>>>

A system is deterministic if, given the system's complete state at a point in
time, it is possible, in principle, to predict its future behavior
perfectly.  We say ``in principle'' because in the real world,
prediction may be possible but hopelessly impractical.  We write
complex software systems in many cases because we cannot predict their
behavior (if we could perform the calculations in advance, ourselves,
we would just do so).  As a consequence, in software, rather than
defining determinism in terms of prediction, we usually therefore
simply say that a system is deterministic if, given the same state and
inputs, it always produces the same outputs.

@@

Because the
programmer's abstraction of the system ignores  such a large number of
details, very few tests are in fact ``deterministic'' in the sense
that they eliminate all changes in behavior between executions.  Running the same test
twice almost always results in differences, given a low enough level of
abstraction, since the system load,
cache contents, branch predictor history, etc. are almost never
controlled for; however, this kind of
nondeterminism is usually of no interest, unless the test involves
extremely precise timing constraints.  Rather, what matters is when
some kind of nondeterminism unexpectedly impacts computed values at a
higher level of abstraction; in general, if the operating system
itself is not buggy, low-level nondeterminism, by design, is invisible
except in fine-grained performance testing or real-time systems.
Unexpected nondeterminism usually arises when there is an element of higher-level
state or input that \emph{is} critical to the produced behavior, but
the programmer has not anticipated.  E.g.,  when it is believed that the
behavior of a thread scheduler will not matter, but a race condition
in the code means that it does matter, or when the order of items in
an iterator on a hash table is important, and the hash values used are
randomly salted.

@@

In order to help ensure that they are reliable and secure, complex
modern software systems usually include a large set of
\emph{regression tests}.  A regression test suite is a (usually large)
set of tests that can be run against a software system every time it
is modified, to ensure that the modification has not broken the system
in some way.  \emph{Flaky tests} \cite{miccoflaky} are regression
tests that fail in an intermittent, unreliable fashion, and thus
degrade the utility of regression testing.  The essence of a flaky
test is that, for the same snapshot of test code and code under test
(CUT), it sometimes fails and sometimes passes: the pass/fail result (disposition) of the
test is not a deterministic property of the test code, code under
test, and environment running the test.  This produces three serious
problems: first, a flaky test often wastes developer time and delays
software changes by forcing the investigation of code-under-test that
is not actually incorrect.  Second, failures in flaky tests (for that
reason) are often ignored, and therefore serious software faults may
be missed.  Finally, to mitigate these problems, flaky tests are often
run multiple times, wasting valuable computing resources and also
delaying acceptance of code changes.  An analogy may make the extent
of the problem in large-scale test automation clear:  a canary in a coal mine is of
little use if canaries frequently become ill for reasons unrelated to
the presence of toxic gases.  Mining may stop for no good reason, or
miners may learn to ignore the canary, leading to tragedy; a third,
more ``practical'' option is that miners may carry so many redundant
canaries into the coal mine that canary-care becomes a serious burden
on mining.

@@

This paper proposes (1) a number of formal definitions of types of
nondeterminism and (2) an implementation, based on these definitions, for detecting and debugging
nondeterminism in property-based testing.  This enables a number of useful results:
 the automated production of
nondeterminism-free regression tests even for libraries with some
sources of nondeterminism, the documentation of
nondeterminism, and, most importantly, detection and effective debugging for
unexpected nondeterminism that is, itself, a bug.  We propose
modifications to the widely used delta-debugging algorithm in order to better handle
nondeterminism as a test property, and describe an implementation of
our approach in a Python testing tool that is used on real-world
projects.  Finally, we use a set
of case studies of real, widely-used Python projects to demonstrate the
utility of our ideas.
